CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-05-26 script/en/chapters/digitalsignatures.tex
be updates
*** empty log message ***

\newpage
\section{Digital Signatures}
\index{Signatures!digital}
(Schneider J. / Esslinger B. / Filipovics B. / Koy H., June 2002,
Update: Feb. 2003)

@@

In detail, a \index{Signature procedure} {\em Signature procedure} looks like
this: \\ Senders use their message and secret key to calculate the digital
signature for the message. Compared to hand-written signatures, digital
signatures therfore have the advantage that they also depend on the document to
be signed. Signatures from one and the same participant are different unless the
signed documents are completely identical. Even inserting a blank in the text
would lead to a different signature. The recipient of the message would
therefore detect any injury to the message integrity as this would mean that the
signature no longer matches the document and is shown to be incorrect when
verified.

The document is sent to the recipient together with the signature. The recipient
can then use the sender's public key, the document and the signature to
establish whether or not the signature is correct. In practice, however, the
procedure we have just described has a decisive disadvantage. The signature is
approximately as long as the document itself. To prevent an unnecessary increase
in data traffic, and also for reasons of performance\index{Performance}, we use 
a cryptographic hash function\footnote{%
%xxxxxxxxxx Menüeinträgsbezeichnungen ok?
Hash functions\index{Hash function} are implemented within CrypTool\index{CrypTool} v1.3.04 at
several places. \\
Using menu {\bf Indiv. Procedures \textbackslash{} Hash} you have the possibilities
% hier die items nicht einrücken!
\begin{list}{\textbullet}{\leftmargin10pt\addtolength{\itemsep}{-1.0\baselineskip}}
%\begin{itemize}\addtolength{\itemsep}{-1.0\baselineskip} % um den Abstand zu verringern bebe
\item to apply one of 6 hash functions to the content of the current window, \\
\item to calculate the hash value of a file, \\
\item to test, how changes to a text change the according hash value and \\
\item to perform a simulation, how digital signatures could be attacked 
      by a targeted search for hash value collisions.
\end{list}
}.

A cryptographic \index{Hash function} {\em hash function} maps a message of any
length to a string of characters with a constant size (usually 128 or 160 bits),
the \index{Hash value} hash value. It should be practically impossible, for a
given number, to find a message that has precisely this number as hash value.
Furthermore, it should be practically impossible to find two messages with the
same hash value. In both cases the final signature procedure would display weak
points.

So far, no formal proof has been found that perfectly secure cryptographic hash
functions exist. However, there are several good candidates that have not yet
shown any weak points in practice (e.g. \index{SHA-1} SHA-1 or \index{RIPEMD-160} RIPEMD-­160).

@@

\subsection{RSA signatures}
\index{Signatures!digital}
\index{RSA signatures}

@@

\subsection{DSA signatures}
\index{Signatures!digital}
\index{DSA signatures}

@@

\emph{Remark:}
\begin{itemize}
\item $(s,r)$ is the signature.
\item The security of the signature depends not only on the mathematical
  properties, but also on using a good random source\index{Random}  for $k$.
\item SHA-1 \index{SHA-1} is a 160-bit hash function specified in FIPS186.
\end{itemize}
\paragraph{Verifying}\strut\\
\begin{tabular}{l@{ }l}
$w$ & $= s^{-1}  \mbox{ mod } q$\\
$u_1$ & $= (\mbox{SHA-1}(m)w) \mbox{ mod } q$\\
$u_2$ & $= (rw)  \mbox{ mod } q$\\
$v$ & $= (g^{u_1}y^{u_2}) \mbox{ mod } p)  \mbox{ mod } q$\\

@@

\subsection{Public key certification}
\index{Certification!public key}
The aim of public key certification is to guarantee the connection between a
public key and a user and to make it traceable for external parties. In cases in
which it is impossible to ensure that a public key really belongs to a
particular person, many protocols are no longer secure, even if the individual
cryptographic modules cannot be broken.

\subsubsection{Impersonation attacks}
\index{Impersonation attack}
Assume Charlie has two pairs of keys (PK1, SK1) and (PK2, SK2), where SK denotes
the secret key and PK the public key. Further assume that he manages to palm off
PK1 on Alice as Bob's public key and PK2 on Bob as Alice's public key (by
falsifying a public key directory).

@@

\subsubsection{X.509}

All participants who want to \index{X.509} have a X.509 certificate verifying that their public
key belongs to a real person consult what is known as a
\index{Certification authority (CA)} certification authority (CA). They prove their identity to this
CA (for example by showing their ID). The CA then issues them an electronic
document (certificate) which essentially contains the names of the certificate-holder
and the CA, the certificate-holder's public key and the validity period
of the certificate. The CA then signs the certificate using its secret key.

@@

\bibitem[FIPS186]{FIPS186} U.S. Department of Commerce/N.I.S.T.
    \index{FIPS186} \\
    Entity authentication using public key cryptography, February 18, 1997.

>>>>

\newpage
% --------------------------------------------------------------------------
\section{Digital Signatures}
\index{Signature!digital}
(Schneider J. / Esslinger B. / Filipovics B. / Koy H., June 2002,
Update: Feb. 2003)

@@

In detail, a \index{Signature procedure} {\em Signature procedure} looks like
this: \\ Senders use their message and secret key to calculate the digital
signature for the message. Compared to hand-written signatures, digital
signatures therefore have the advantage that they also depend on the document to
be signed. Signatures from one and the same participant are different unless the
signed documents are completely identical. Even inserting a blank in the text
would lead to a different signature. The recipient of the message would
therefore detect any injury to the message integrity as this would mean that the
signature no longer matches the document and is shown to be incorrect when
verified.

The document is sent to the recipient together with the signature. The recipient
can then use the sender's public key, the document and the signature to
establish whether or not the signature is correct. In practice, however, the
procedure we have just described has a decisive disadvantage. The signature is
approximately as long as the document itself. To prevent an unnecessary increase
in data traffic, and also for reasons of performance\index{Performance}, we use 
a cryptographic hash function\footnote{%
%xxxxxxxxxx Menüeinträgsbezeichnungen ok?
Hash functions\index{Hash function} are implemented within CrypTool\index{CrypTool} v1.3.04 at
several places. \\
Using menu {\bf Indiv. Procedures \textbackslash{} Hash} you have the possibilities
% hier die items nicht einrücken!
\begin{list}{\textbullet}{\leftmargin10pt\addtolength{\itemsep}{-1.0\baselineskip}}
%\begin{itemize}\addtolength{\itemsep}{-1.0\baselineskip} % um den Abstand zu verringern bebe
\item to apply one of 6 hash functions to the content of the current window, \\
\item to calculate the hash value of a file, \\
\item to test, how changes to a text change the according hash value and \\
\item to perform a simulation, how digital signatures could be attacked 
      by a targeted search for hash value collisions\index{collisions}.
\end{list}
}.

% --------------------------------------------------------------------------
\subsection{Hash functions}
\index{Hash function}
A cryptographic \index{Hash function} {\em hash function}\footnote{%
Secure hash algorithms compute a condensed representation of 
electronic data (message).
When a message is input to a hash algorithm, the result is an output called
a message digest. The message digests typically range in length 
from 128 to 512 bits, depending on the algorithm. 
Secure hash algorithms are typically used with
other cryptographic algorithms, such as digital signature algorithms and
keyed-hash message authentication codes, or in the generation of random
numbers (bits).
}
maps a message of any
length to a string of characters with a constant size,
the \index{Hash value} hash value. It should be practically impossible, for a
given number, to find a message that has precisely this number as hash value.
Furthermore, it should be practically impossible to find two messages with the
same hash value\index{collisions}. 
In both cases the final signature procedure would include weaknesses.

So far, no formal proof has been found that perfectly secure cryptographic
hash functions exist. However, there are several good candidates that have
not yet shown any weak points in their structure in practice 
(e.g. \index{SHA-1} SHA--1\footnote{%
SHA-1 \index{SHA-1} is a 160 bit hashfunction specified in FIPS 180-1, 
ANSI X9.30 Part 2 and \cite{FIPS186}.\\
SHA means Secure Hash Algorithm, and is widely used, 
e.g. with DSA, RSA or ECDSA.\\
The current standard \cite{FIPS180-2} defines four secure hash algorithms
-- SHA-1, SHA-256, SHA-384, and SHA-512.

For these hash algorithms there are also validation tests defined in the
test suite FIPS 140-2.

The output length of the SHA algorithms was enhanced because of the
possibility of birthday attacks\index{Birthday attack}\index{collisions}:
these make n-bit AES and a 2n-bit hash roughly equivalent: \\
- 128-bit AES -- SHA-256 \\
- 192-bit AES -- SHA-384 \\
- 256-bit AES -- SHA-512.

With CrypTool\index{CrypTool} v1.3.04 you can comprehend the birthday attack
\index{Birthday attack} on digital signatures: \\
using menu {\bf Indiv. Procedures \textbackslash{} 
Attack on the hashvalue of the digital signature}.

}  or \index{RIPEMD-160} RIPEMD-­160\footnote{%
RIPEMD-160, RIPEMD-128 and the optional extension RIPEMD-256 have object
identifiers defined by the ISO-identified organization TeleTrusT, both 
as hash algorithm and in combination with RSA.

RIPEMD-160 is also part of the ISO/IEC international standard 
ISO/IEC 10118-3:1998 on dedicated hash functions, together with 
RIPEMD-128 and SHA-1.

Further details: \\
http://www.esat.kuleuven.ac.be/~bosselae/ripemd160.html \\
http://www.ietf.org/rfc/rfc2857.txt ("The Use of HMAC-RIPEMD-160-96 within ESP and AH").
}
).

@@

% --------------------------------------------------------------------------
\subsection{RSA signatures}
\index{Signature!digital}
\index{RSA!signature}

@@

% --------------------------------------------------------------------------
\subsection{DSA signatures}
\index{Signature!digital}
\index{DSA!signature}

@@

\emph{Remark:}
\begin{itemize}
\item $(s,r)$ is the signature.
\item The security of the signature depends not only on the mathematical
  properties, but also on using a good random source\index{Random}  for $k$.
\item SHA-1 \index{SHA-1} is a 160-bit hash function.
\end{itemize}
\paragraph{Verifying}\strut\\
\begin{tabular}{l@{ }l}
$w$ & $= s^{-1}  \mbox{ mod } q$\\
$u_1$ & $= (\mbox{SHA-1}(m)w) \mbox{ mod } q$\\
$u_2$ & $= (rw)  \mbox{ mod } q$\\
$v$ & $= (g^{u_1}y^{u_2}) \mbox{ mod } p)  \mbox{ mod } q$\\

@@

% --------------------------------------------------------------------------
\subsection{Public key certification}
\index{Certification!public key}
The aim of public key certification is to guarantee the connection between a
public key and a user and to make it traceable for external parties. In cases in
which it is impossible to ensure that a public key really belongs to a
particular person, many protocols are no longer secure, even if the individual
cryptographic modules cannot be broken.

% --------------------------------------------------------------------------
\subsubsection{Impersonation attacks}
\index{Impersonation attack}
Assume Charlie has two pairs of keys (PK1, SK1) and (PK2, SK2), where SK denotes
the secret key and PK the public key. Further assume that he manages to palm off
PK1 on Alice as Bob's public key and PK2 on Bob as Alice's public key (by
falsifying a public key directory).

@@

% --------------------------------------------------------------------------
\subsubsection{X.509}
All participants who want to \index{X.509} have a X.509 certificate verifying that their public
key belongs to a real person consult what is known as a
\index{Certification authority (CA)} certification authority (CA). They prove their identity to this
CA (for example by showing their ID). The CA then issues them an electronic
document (certificate) which essentially contains the names of the certificate-holder
and the CA, the certificate-holder's public key and the validity period
of the certificate. The CA then signs the certificate using its secret key.

@@

\bibitem[FIPS180-2]{FIPS180-2} U.S. Department of Commerce/N.I.S.T. 
    \index{FIPS180-2} \\
    Secure Hash Standard (SHS), August 1, 2002.\\
    (FIPS 180-2 will supersede FIPS 180-1 beginning February 1, 2003.)

\bibitem[FIPS186]{FIPS186} U.S. Department of Commerce/N.I.S.T.
    \index{FIPS186} \\
    Entity authentication using public key cryptography, February 18, 1997.\\
    No more valid.

\bibitem[FIPS186-2]{FIPS186-2} U.S. Department of Commerce/N.I.S.T. 
    \index{FIPS186-2} \\
    Digital Signature Standard (DSS), January 27, 2000.\\
    Change Note: October 5, 2001.\\
    http://csrc.nist.gov/publications/fips/fips186-2/fips186-2-change1.pdf

### 2003-05-26 script/en/chapters/introduction.tex
*** empty log message ***

Chapter 6 of this script has also been enhanced: now we explain in greater
detail how elliptic curves are defined over different number fields.

>>>>

Chapter 5 (more about hash algorithms) and chapter 6 (now we explain in
greater detail how elliptic curves are defined over different number
fields) of this script have also been enhanced.

### 2003-05-26 script/en/chapters/menus.tex
*** empty log message ***

\pagebreak
\enlargethispage{1cm}
\section{CrypTool Menus}
\hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
This appendix contains the complete menu tree of CrypTool. 
Please notice that not all menu items are always available, because the
menu tree is built up dynamically: which menu items are available depends
on the type of the currently active document window.
The brute-force analysis for DES e.~g. is only available, if the active
window is opened in the hexadecimal view. On the other hand the menu item
"`Generate Random Numbers\dots"' is always available. 
Within the figures \ref{menu-detail-1} and \ref{menu-detail-2} 
this relation is shown in the brackets after the menu items,
e.~g. "`Find\dots[ATH]"' means, that the function "`Find\dots"' is only
available for ASC-, text- and hexadecimal windows.
\begin{center}
\begin{tabular}{rl}
\bf Code letter & \bf Type of document \\
M & (no document opened)\\
A & ASC\\
T & Text\\
H & Hexadecimal\\
P & Plot\\
\end{tabular}
\end{center}

@@

\begin{figure}[!hb]
\begin{center}
\includegraphics[scale=1.1, clip, viewport=10 360 400 600]{figures/cryptool-menu-detail-en}
\caption{Detailled view of the CrypTool menu tree -- part 1}
\label{menu-detail-1}
\end{center}
\end{figure}

\begin{figure}[b]
\begin{center}
\includegraphics[scale=1.1, clip, viewport=400 190 784 598]{figures/cryptool-menu-detail-en}
\caption{Detailled view of the CrypTool menu tree -- part 2}
\label{menu-detail-2}
\end{center}
\end{figure}

\begin{figure}[hb]
\begin{center}
\includegraphics[scale=0.75, angle=270, viewport=14 107 779 590]{figures/cryptool-menu-en}
\vspace{-18pt}
\caption{Complete Overview of the CrypTool menu tree} 
\label{menuoverview}
\end{center}
\end{figure}

>>>>

\pagebreak
\enlargethispage{1cm}
\section{CrypTool Menus}
\hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
This appendix contains the complete menu tree of CrypTool. 
Please notice that not all menu items are always available, because the
menu tree is built up dynamically: which menu items are available depends
on the type of the currently active document window.
The brute-force analysis for DES e.~g. is only available, if the active
window is opened in the hexadecimal view. On the other hand the menu item
``Generate Random Numbers\dots'' is always available. 
Within the figures \ref{menu-detail-1} and \ref{menu-detail-2} 
this relation is shown in the brackets after the menu items,
e.~g. ``Find\dots[ATH]'' means, that the function ``Find\dots'' is only
available for ASC-, text- and hexadecimal windows.
\begin{center}
\begin{tabular}{rl}
\bf Code letter & \bf Type of document \\
M & (no document opened)\\
A & ASC\\
T & Text\\
H & Hexadecimal\\
P & Plot\\
\end{tabular}
\end{center}

@@

\begin{figure}[!hb]
\begin{center}
%\frame{
\includegraphics[scale=1.1, clip, viewport=0 300 404 591]{figures/cryptool-menu-detail-en}
%}
\caption{Detailed view of the CrypTool menu tree -- part 1}
\label{menu-detail-1}
\end{center}
\end{figure}

\begin{figure}[b]
\begin{center}
%\frame{
\includegraphics[scale=1.1, clip, viewport=404 117 828 598]{figures/cryptool-menu-detail-en}
%}
\caption{Detailed view of the CrypTool menu tree -- part 2}
\label{menu-detail-2}
\end{center}
\end{figure}
\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
%\frame{
\includegraphics[scale=0.75, angle=270, viewport=14 5 829 590]{figures/cryptool-menu-en}
%}
\caption{Complete overview of the CrypTool menu tree} 
\label{menuoverview}
\end{center}
\end{figure}
\clearpage

### 2003-05-26 script/en/chapters/moderncryptography.tex
*** empty log message ***

In this process, we describe a problem as ``easy'' if it can be solved in 
\index{Runtime!polynomial} \index{Polynomial }polynomial time as a function of the length of the 
input. If the length of the input is $ n $ bits, then the time for calculating 
the function is proportional to $ n^{a}, $ where $ a $  is a constant. We say 
that the \index{Complexity} complexity of such problems is $ O(n^{a}) $
[Landau- or Big-O notation].

@@

In this case it has been shown that the procedure \index{Rabin} 
\index{Rabin!public key} is equivalent to breaking the factorisation problem. 
Unfortunately, this procedure is susceptible to chosen-cipher text attacks.
\index{Attack!chosen-cipher text}
\begin{center}
\fbox{\parbox{12cm}{        
Let $ p $ and $ q $ be two different prime numbers with $ p,q\equiv 3 \mod 4 $ 
and $ n = pq.$ Let $ 0\leq B \leq n-1.$ \\
\underline{Public key:}
\[ e=(n,B). \]
\underline{Private key:}
\[ d=(p,q). \]
\underline{Encryption:}
\[ y= e_{T} (x) = x(x+B) \mod n.\]
\underline{Decryption:}
\[ d_{T} (y) = \sqrt{y + B^2/4} -B/2 \mod n. \]
}}
\end{center}

>>>>

In this process, we describe a problem as ``easy'' if it can be solved in 
\index{Runtime!polynomial} \index{Polynomial}polynomial time as a function of the length of the 
input. If the length of the input is $ n $ bits, then the time for calculating 
the function is proportional to $ n^{a}, $ where $ a $  is a constant. We say 
that the \index{Complexity} complexity of such problems is $ O(n^{a}) $
[Landau- or Big-O notation].

@@

In this case it has been shown that the procedure \index{Rabin Michael O.} 
\index{Rabin!public key procedure} is equivalent to breaking the factorisation problem. 
Unfortunately, this procedure is susceptible to chosen-cipher text attacks.
\index{Attack!chosen-cipher text}
\begin{center}
\fbox{\parbox{12cm}{        
Let $ p $ and $ q $ be two different prime numbers with $ p,q\equiv 3 \mod 4 $ 
and $ n = pq.$ Let $ 0\leq B \leq n-1.$ \\
\underline{Public key:}
\[ e=(n,B). \]
\underline{Private key:}
\[ d=(p,q). \]
\underline{Encryption:}
\[ y= e_{T} (x) = x(x+B) \mod n.\]
\underline{Decryption:}
\[ d_{T} (y) = \sqrt{y + B^2/4} -B/2 \mod n. \]
}}
\end{center}
